---
title: "Scraping WW2 casualties from Wikipedia"
author: "Tina Maria Villesen Skov and Mie Højfeldt Kristensen"
date: "4 Dec 2021, updated `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal
Our goal is to make a visualization of which countries suffered the biggest human loses during WW2.

# Challenge
We found the data on https://en.wikipedia.org/wiki/World_War_II_casualties . We then need to srape that particular html table and use that to make our visualization.

# Solution
First, install a handful of classic R packages and load their libraries:

- `rvest` for web-scraping
- `dplyr` for data-wrangling
- `tidyr` for data transformation
- `stringr` for string manipulation
- `janitor` for clean headers that your OCD will love you for

```{r libraries, warning=FALSE, message=FALSE}
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(varhandle)
library(ggplot2)
```

## Scrape the data
We scrape the html table with rvest library.
```{r}
# scrape the website
pageHtml <- read_html("https://en.wikipedia.org/wiki/World_War_II_casualties")

# Read whole html table and extract first table that holds the data we want to work with.
casualtiesTable <- pageHtml %>%
 html_nodes("table") %>%
 html_table()
casualtiesTable <- casualtiesTable[[1]]
head(casualtiesTable)
```

## Cleaning scraped data

### Rename columns
Column names in the original table are long. We shorten them for easier referencing them in code and make the code more readable.
```{r wrangle-columns, message = FALSE, warning=FALSE}
library(tidyverse)
casualtiesTable <- as_tibble(casualtiesTable, .name_repair = c("universal"))

data <- casualtiesTable %>%
	rename(TotalPop1939 = "Total.population1.1.1939") %>%
  rename(MilitaryDeaths = "Militarydeaths.from.all.causes") %>%
  rename(CivilianDeathsDueToMilitary = "Civilian.deaths.due.tomilitary.activity.and.crimes.against.humanity") %>%
  rename(CivilianDeathsDueToDiseaseFamine = "Civilian.deaths.due.towar.related.famine.and.disease") %>%
  rename(TotalDeaths = "Totaldeaths") %>%
  rename(DeathPctOf1939Pop = "Deaths.as...of1939.population") %>%
  rename(AverageDeathPctOf1939Pop = "Average.Deaths.as...of1939.population")
```

### Adjust the data
In this section we have created several helper function to fix our scrapped data.
Data from Wikipedia has footnotes appended to most of the texts and values. So thoses are removed.
Numeric values are comma formatted so those are removed and we cast value to numertic data type.

The different helper functions has comments in each that explains what goes on.
```{r, warning=FALSE, message=FALSE}

# Helper function to remove footnote uppercase character that was scrapped from Wiki
removeFootnoteChar <- function(x){
  lastChar <- substr(x, nchar(x), nchar(x))
 
  while(str_detect(lastChar,"[[:upper:]]")){
    x <- substr(x, 0, nchar(x)-1)
    lastChar <- substr(x, nchar(x), nchar(x))
  }
  return(x)
}

# Helper to get raw numeric value from formatted string containing numeric value
getNumericFromString <- function(x){
  # If string contains "to" do and avg of two values. Eg "2,400,000[39] to 4,000,000[40]"
  if(str_detect(x, "to")){
    numList <- strsplit(x, " to ")
    numList <- numList[[1]]
    sum <- 0
    for(i in numList){
      i<-gsub("\\[.*","",i)
      i<-gsub(",", "", i)
      i<-gsub("\\+", "", i)
      
      # Some values has footnote appended. Remove them 
      if(str_detect(i,"[[:upper:]]"))
        i <- removeFootnoteFromNumeric(i)
      
      sum = sum + as.numeric(i)
    }
    return(sum / length(numList))
  } else{
    x<-gsub("\\[.*","",x)
    x<-gsub(",", "", x) 
    x<-gsub("\\+", "", x)
    
    # Some values has footnote appended. Remove them 
    if(str_detect(x,"[[:upper:]]"))
        x <- removeFootnoteFromNumeric(x)
    
    return(as.numeric(x))
  }
}

# Helper to remove footnotes from numeric strings
removeFootnoteFromNumeric <- function(x){
  strIndex <- 1;
  result <- ""
  # Find first non numeric value and cut it form the numeric value
  while(nchar(x) >= strIndex){
    result <- substr(x, 1, strIndex)
    
    # Using varhandler lib we can use the check_numeric function, and detect digits inside string   
    if(!check.numeric(result)){
      result <- substr(x, 1, strIndex-1)
      break
    }
    
    # Increment counter
    strIndex <- strIndex + 1
  }
  
  return(result)
}

# Clanup data
data <- data %>%
  # Operate the below functions on each single row
  rowwise() %>%
  # Clean country names from footnote chars
	mutate(Country = removeFootnoteChar(Country)) %>%
  # Clean numeric values
  mutate(TotalPop1939 = getNumericFromString(TotalPop1939)) %>%
  mutate(MilitaryDeaths = getNumericFromString(MilitaryDeaths)) %>%
  mutate(CivilianDeathsDueToMilitary = getNumericFromString(CivilianDeathsDueToMilitary)) %>%
  mutate(CivilianDeathsDueToDiseaseFamine = getNumericFromString(CivilianDeathsDueToDiseaseFamine)) %>%
  mutate(TotalDeaths = getNumericFromString(TotalDeaths))
  #mutate(Militarywounded = getNumericFromString(Militarywounded))
```

### Create and read csv
The ggplot library takes a list of data frames. So we first create a csv file from our html scrap and then reads it out as data frames.
```{r write-to-csv, warning=FALSE, message=FALSE}
write_csv(data,"data/ww2_casualties.csv")
data <- read_csv("data/ww2_casualties.csv")
```

## Who suffered most loses?
Now we are ready to plot our graph with our scrapped data. We could do lots of different graphs and comparisons but due to time and scope of this assignment we have endup illustrating military and civilian loses. We wanted to a faction comparioson as well but that is for another time.

### Military loses
We only show countries with more than 250.000 deaths to make the graph more clear. We could have done facet with all the countries in but we want to see the biggest loses.
```{r fig.width = 14, warning=FALSE, message=FALSE}

data %>%
  filter(MilitaryDeaths >= 250000) %>% 
  ggplot(aes(x = Country, y = MilitaryDeaths, fill = Country)) +
  geom_bar(stat='identity') +
  scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
  scale_y_continuous(labels=function(n){format(n, scientific = FALSE, big.mark=",")}) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
  labs(y = "Military deaths",
       title = "More than 250.000 military loses by country") +
  theme(legend.position="bottom")
```

### Civilian loses
Here we compare civilian loses by country and by if the death was a result of military actions or diseas/famine
```{r fig.width = 14, warning=FALSE, message=FALSE}
# Transform data from wide table to long table
data_long <- gather(data, deaths, measurement, CivilianDeathsDueToMilitary:CivilianDeathsDueToDiseaseFamine, factor_key=TRUE)
  ggplot(data_long, aes(x=Country, y=measurement, fill=deaths)) +
  geom_bar(stat='identity', position='dodge') +
  scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
  scale_y_continuous(labels=function(n){format(n, scientific = FALSE, big.mark=",")}) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
  labs(y = "Civilian deaths",
       title = "Civilian loses by country") +
  theme(legend.position="bottom")
```

### Total loses
Visualizing total loses for countries with more than 250.000 deaths
```{r fig.width = 14, warning=FALSE, message=FALSE}
# Pie chart
data %>%
  filter(TotalDeaths >= 250000) %>% 
  ggplot(aes(x="", y=TotalDeaths, fill=Country)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_y_continuous(labels=function(n){format(n, scientific = FALSE, big.mark=",")}) +
  coord_polar("y", start=0) +
  theme(legend.position="bottom") +
  theme_void() 

# Bar chart
data %>%
  filter(TotalDeaths >= 250000) %>% 
  ggplot(aes(x = Country, y = TotalDeaths, fill = Country)) +
  geom_bar(stat='identity') +
  scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
  scale_y_continuous(labels=function(n){format(n, scientific = FALSE, big.mark=",")}) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
  labs(y = "Total deaths",
       title = "Total deaths") +
  theme(legend.position="none")
```
### Conclussion
So stating the obvious. The Soviet Union had drasticaly more deaths compared to the rest. Even comparing to the rest of the allied forces. It has a quite powerfull effect when it is visualized. A bit of an eye opener is that China also had huge civilian loses. In the western world we are not that familiar with what actually happend in east asia during WW2. 